Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-05-16 Text/migration.tex
First go-over after a long dormant period, up to section 3.
	modified:   Text/migration.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@75 3b7f7548-5c66-0410-83ab-3997167635fc

\usepackage{graphicx}
\citestyle{aa}
\usepackage{natbib}

@@

\author{Aaron C. Boley}
\affil{Department of Astronomy; University of Florida,
211 Bryant Space Science Center,
PO Box 112055, Gainesville, FL 32611}
\email{aaron.boley@gmail.com}

\begin{abstract}
%Meaningless Change
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet interacts with a strong spiral mode, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.} 
%Scott: Fill in some references you know now. Aaron: Suggestions on references?
Migration of gas giant planets due to interactions with a circumstellar gas disk probably plays a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However,  the self-gravity of the gas disk has usually been ignored, and the effects of radiative cooling and a detailed equation of state (EOS) have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2010b}. \citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where this probably should not realistically have happened \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boleydurisen2010}. The occurrence of gravitational instabilities (GIs) may also be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

{\bf See note in Latex file.}
%Scott and Aaron: Verify that everything I say here is correct.
The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code \citep{boley2007b} is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ \citep{boley2007} and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a grid size ($\varpi$,$\phi$,z) = (512,512,64). 
{\bf See note in Latex file.} 
%Scott: The resolution stays the same. The grid size varies, but this is unimportant because initially the outer 256 radial zones are empty. It is simplest to just give the eventual grid size. So I did change "resolution" to grid size here.
The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in \citet{boley2007b}, where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in \citet{michael2010}. The planet integration is done with a Verlet integrator \citep[e.g.,][]{hut1995}, and the indirect potential terms are treated as in \citet{nelson2000a}. The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in \citet{boley2006,boley2007b}, except that we correct the \citet{dalessio2001} gas mean molecular weight to $\mu$ = 2.39.

@@

The model disk, based on an equlibrium model in \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 250 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of ??? at radial grid zone ??? (??? AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. There is an inner cylindrical hole in the grid with a radius of 12 radial zones {\bf Scott Comment: brief description of hole needed. Dick's reply in Latex file}. 
%Scott: So add one!
To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, goes from time $t = 0$ to ??? 20 ??? ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10$ ORP. These three simulations are part of a much larger suite, to be described elsewhere, in which the planet mass and starting locations are varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using at the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, attained by the planet over each interval that the planet's $\phi$-coordinate changes by 2$\pi$ (one ``orbit'').  Then, for that orbit, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are only indicative; there are strong orbit to orbit variations in $a$ and $e$.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The red and black indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The diamonds are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planete.pdf}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The red and black indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

>>>>

\usepackage{graphicx}
\citestyle{aa}
\usepackage{natbib}
\usepackage{epstopdf}
\usepackage{grffile}

@@

\author{Aaron C. Boley}
\affil{Department of Astronomy University of Florida, Gainesville, FL 32611}
\email{aaron.boley@gmail.com}

\begin{abstract}
%Meaningless Change
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet interacts with a strong spiral mode, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file. SM: confirmed from radial plots}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.} 
%Scott: Fill in some references you know now. Aaron: Suggestions on references?
Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However,  the self-gravity of the gas disk has usually been ignored, and the effects of radiative cooling and a detailed equation of state (EOS) have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2010b}{\bf NEED TO RECHECK PAARDEKOOPER FOR RADIATIVE COOLING}. \citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where, in reality, this probably would not have happened  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boleydurisen2010}. The occurrence of gravitational instabilities (GIs) may also be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code \citep{boley2007b} is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ \citep{boley2007} and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a grid size ($\varpi$,$\phi$,z) = (512,512,64). 
The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in \citet{boley2007b}, where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in \citet{michael2010}. The planet integration is done with a Verlet integrator \citep[e.g.,][]{hut1995}, and the indirect potential terms are treated as in \citet{nelson2000a}. The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in \citet{boley2006,boley2007b}, except that we correct the \citet{dalessio2001} gas mean molecular weight to $\mu$ = 2.39.

@@

The model disk, based on an equlibrium model in \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of ??? at radial grid zone ??? (??? AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the {\sl fiducial run}, goes from time $t = 0$ to 21 ORP without a planet. In the second simulation, which we call the {\sl t = 0 planet run}, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the {\sl t = 10 planet run}, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10.5$ ORP. These three simulations are part of a much larger suite, to be described elsewhere, in which the planet mass and starting locations are varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by using at the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, attained by the planet over each interval that the planet's $\phi$-coordinate changes by 2$\pi$ (one ``orbit'').  Then, for that orbit, $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These parameters are only indicative; there are strong orbit to orbit variations in $a$ and $e$.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planeta.eps}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The red and black indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The diamonds are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/planete.eps}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The red and black indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}
